\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2020-03-06 analysis/paper/paper.Rmd
Fixed lollipop plot
to do:
fix casebase absoluterisk for glmnet. 
fix ??? tags

---
documentclass: jss
author:
  - name: Sahir Bhatnagar *
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{sahir.bhatnagar@mail.mcgill.ca}
    url: http://sahirbhatnagar.com/
  - name: Maxime Turgeon *
    affiliation: 'University of Manitoba \AND'
    address: >
      186 Dysart Road
      Winnipeg, MB, Canada R3T 2N2
    email: \email{max.turgeon@umanitoba.ca}
    url: https://maxturgeon.ca/
  - name: Jesse Islam 
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{jesse.islam@mail.mcgill.ca}
  - name: James Hanley
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{james.hanley@mcgill.ca}
    url: http://www.medicine.mcgill.ca/epidemiology/hanley/
  - name: Olli Saarela
    affiliation: University of Toronto
    address: >
      Dalla Lana School of Public Health, 
      155 College Street, 6th floor, 
      Toronto, Ontario 
      M5T 3M7, Canada
    email: \email{olli.saarela@utoronto.ca}
    url: http://individual.utoronto.ca/osaarela/
title:
  formatted: "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
  # If you use tex in the formatted title, also supply version without
  plain:     "casebase: An Alternative Framework For Survival Analysis"
  # For running headers, if needed
  short:     "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
abstract: >
  The abstract of the article. * joint co-authors
keywords:
  # at least one keyword must be supplied
  formatted: [keywords, not capitalized, "\\proglang{Java}"]
  plain:     [keywords, not capitalized, Java]
preamble: >
  \usepackage{amsmath}
  \usepackage{longtable}
  \usepackage{graphicx}
  \usepackage{tabularx}
  \usepackage{float}
  \usepackage{booktabs}
  \usepackage{makecell}
  \DeclareUnicodeCharacter{2500}{-}
output: 
  rticles::jss_article:
    number_sections: TRUE     #added argument option 
    citation_package: "natbib"  #All my citations use biblatex, not natbib. 
biblio-style: jss      #Listed to use in JSS Instructions for Authors, but not in template by default. 
bibliography: references.bib  #Also not included in template by default. 
fig_width: 7
fig_height: 6
fig_caption: true
---

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = TRUE,
  comment = "#>",
  fig.path = "../figures/"
)
options(kableExtra.latex.load_packages = FALSE)
library(tidyverse)
library(magrittr)
library(survival)
library(casebase)
#devtools::load_all('../../../casebase')#testing
library(splines)
library(tibble)
library(glmnet)
library(data.table)
library(kableExtra)
library(pracma)
options(digits = 2, scipen = 999)
# set the seed for reproducible output
set.seed(1234)

eval_introduction <- TRUE
eval_theory <- TRUE
eval_implementation <- TRUE
eval_cs1 <- TRUE
eval_cs2 <- TRUE
eval_cs3 <- FALSE
eval_cs4 <- TRUE
```

@@

```{r erspc-data, eval = eval_cs1}
data(ERSPC)
ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
                       levels = c(0, 1), 
                       labels = c("Control group", "Screening group"))
```

@@

```{r plot-stratified-erspc-data, eval = eval_cs1}
pt_object_strat <- casebase::popTime(ERSPC, 
                                     event = "DeadOfPrCa", 
                                     exposure = "ScrArm")
plot(pt_object_strat)
```

@@

```{r erspc-cox, eval = eval_cs1}
cox_model <- survival::coxph(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
                             data = ERSPC)
summary(cox_model)
```

@@

```{r erspc-cox-cif, eval = eval_cs1}
new_data <- data.frame(ScrArm = c("Control group", "Screening group"),
                       ignore = 99)

plot(survfit(cox_model, newdata = new_data),
     xlab = "Years since Randomization", 
     ylab = "Cumulative Incidence (%)", 
     fun = "event",
     xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
     main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
                    exp(coef(cox_model)), 
                    exp(confint(cox_model))[1], 
                    exp(confint(cox_model))[2]),
     ylim = c(0, 2), yscale = 100)
legend("topleft", 
       legend = c("Control group", "Screening group"), 
       col = c("red","blue"),
       lty = c(1, 1), 
       bg = "gray90")
```

@@

```{r erspc-casebase-exponential, eval = eval_cs1}
casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm, 
                                                  data = ERSPC, 
                                                  ratio = 100)

summary(casebase_exponential)
exp(coef(casebase_exponential)[2])
exp(confint(casebase_exponential)[2,])
```

@@

```{r erspc-casebase-cif, eval = eval_cs1}
smooth_risk_exp <- casebase::absoluteRisk(object = casebase_exponential, 
                                          time = seq(0,15,0.1), 
                                          newdata = new_data)

plot(survfit(cox_model, newdata = new_data),
     xlab = "Years since Randomization", 
     ylab = "Cumulative Incidence (%)", 
     fun = "event",
     xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
     main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
                    exp(coef(cox_model)), 
                    exp(confint(cox_model))[1], 
                    exp(confint(cox_model))[2]),
     ylim = c(0, 2), yscale = 100)
lines(smooth_risk_exp[,1], smooth_risk_exp[,2], col = "red", lty = 2)
lines(smooth_risk_exp[,1], smooth_risk_exp[,3], col = "blue", lty = 2)

legend("topleft", 
       legend = c("Control group (Cox)","Control group (Casebase)",
                  "Screening group (Cox)", "Screening group (Casebase)"), 
       col = c("red","red", "blue","blue"),
       lty = c(1, 2, 1, 2), 
       bg = "gray90")
```

@@

```{r erspc-casebase-weibull, eval = eval_cs1}
casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm, 
                                 data = ERSPC, 
                                 ratio = 100)

@@

```{r erspc-casebase-weibull-cif, eval = eval_cs1}
smooth_risk_time <- casebase::absoluteRisk(object = casebase_time, 
                                          time = seq(0,15,0.1), 
                                          newdata = new_data)

plot(survfit(cox_model, newdata = new_data),
     xlab = "Years since Randomization", 
     ylab = "Cumulative Incidence (%)", 
     fun = "event",
     xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
     main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
                    exp(coef(cox_model)), 
                    exp(confint(cox_model))[1], 
                    exp(confint(cox_model))[2]),
     ylim = c(0, 2), yscale = 100)
lines(smooth_risk_time[,1], smooth_risk_time[,2], col = "red", lty = 2)
lines(smooth_risk_time[,1], smooth_risk_time[,3], col = "blue", lty = 2)

legend("topleft", 
       legend = c("Control group (Cox)","Control group (Casebase)",
                  "Screening group (Cox)", "Screening group (Casebase)"), 
       col = c("red","red", "blue","blue"),
       lty = c(1, 2, 1, 2), 
       bg = "gray90")
```

@@

```{r erspc-casebase-splines, eval = eval_cs1}
casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
                                    data = ERSPC, 
                                    ratio = 100)

@@

```{r erspc-casebase-splines-cif, eval = eval_cs1}
smooth_risk_splines <- absoluteRisk(object = casebase_splines, 
                                    time = seq(0,15,0.1), 
                                    newdata = new_data)

plot(survfit(cox_model, newdata = new_data),
     xlab = "Years since Randomization", 
     ylab = "Cumulative Incidence (%)", 
     fun = "event",
     xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
     main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
                    exp(coef(cox_model)), 
                    exp(confint(cox_model))[1], 
                    exp(confint(cox_model))[2]),
     ylim = c(0, 2), yscale = 100)
lines(smooth_risk_splines[,1], smooth_risk_splines[,2], col = "red", lty = 2)
lines(smooth_risk_splines[,1], smooth_risk_splines[,3], col = "blue", lty = 2)

legend("topleft", 
       legend = c("Control group (Cox)","Control group (Casebase)",
                  "Screening group (Cox)", "Screening group (Casebase)"), 
       col = c("red","red", "blue","blue"),
       lty = c(1, 2, 1, 2), 
       bg = "gray90")
```

@@

```{r erspc-95ci, echo = FALSE, eval = eval_cs1}
CI_cox <- c(exp(coef(cox_model)), exp(confint(cox_model)))
CI_exp <- c(exp(coef(casebase_exponential))["ScrArmScreening group"], exp(confint(casebase_exponential))["ScrArmScreening group",])
CI_time <- c(exp(coef(casebase_time))["ScrArmScreening group"], exp(confint(casebase_time))["ScrArmScreening group",])
CI_splines <- c(exp(coef(casebase_splines))["ScrArmScreening group"], exp(confint(casebase_splines))["ScrArmScreening group",])

@@

data.frame(Model = c("Cox", "Exponential", "Weibull", "Splines")) %>% 
  bind_cols(as.data.frame(table_ci)) %>% 
  mutate(CI = paste0("(", round(LB, 2), ", ",
                     round(UB, 2), ")"),
         Width = UB - LB) %>% 
  dplyr::select(-LB, -UB) %>% 
  knitr::kable(format = "latex", booktabs = TRUE)  %>%
  kable_styling()
```

@@

```{r erscp-aic, eval = eval_cs1}
c("Exp." = AIC(casebase_exponential),
  "Weib." = AIC(casebase_time),
  "Splines" = AIC(casebase_splines))
```

@@

```{r echo = FALSE, eval = eval_cs2}
tibble::tribble(
    ~Variable, ~Description, ~`Statistical summary`,
    "\\code{Sex}", "Sex", "M=Male (100)",
    "", "", "F=Female (77)",
    "\\code{D}", "Disease", "ALL (73)",
    "", "", "AML (104)",
    "\\code{Phase}", "Phase", "CR1 (47)", 
    "", "", "CR2 (45)", 
    "", "", "CR3 (12)", 
    "", "", "Relapse (73)",
    "\\code{Source}", "Type of transplant", "BM+PB (21)", 
    "", "", "PB (156)",
    "\\code{Age}", "Age of patient (years)", "4--62", 
    "", "", "30.47 (13.04)",
    "\\code{ftime}", "Failure time (months)", "0.13--131.77",
    "", "", "20.28 (30.78)",
    "\\code{Status}", "Status indicator", "0=censored (46)", 
    "", "", "1=relapse (56)", 
    "", "", "2=competing event (75)"
) %>% 
  knitr::kable(format = "latex", booktabs = TRUE,
               caption = "Baseline characteristics of patients in the stem-cell transplant study.", 
               label = "table1bmtcrr",
               escape = FALSE) %>%
  kable_styling()
```

@@

pt_object_comp <- bmtcrr %>% 
    mutate(Status = case_when(Status == 1 ~ 2L, 
                              Status == 2 ~ 1L, 
                              TRUE ~ Status)) %>% 
    casebase::popTime(event = "Status", time = "ftime")

plot_grid(plot(pt_object) + ggtitle("Relapse"), 
          plot(pt_object_comp, point.colour = "blue") +
            ggtitle("Competing events"), 
          labels = c('A', 'B'))
```

@@

```{r bmtcrr-casebase-weibull, warning = FALSE, eval = eval_cs2}
model_cb <- fitSmoothHazard(
    Status ~ ftime + Sex + D + Phase + Source + Age, 
    data = bmtcrr, 
    ratio = 100, 
    time = "ftime")
```

@@

```{r bmtcrr-cox, echo = TRUE, eval = eval_cs2}
library(survival)
# Treat competing event as censoring
model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
                   data = bmtcrr)
```

```{r bmtcrr-cis, echo=FALSE, eval = eval_cs2}
# Table of coefficients----
library(glue)
z_value <- qnorm(0.975)
foo <- summary(model_cb)@coef3
table_cb <- foo[!grepl("Intercept", rownames(foo)) & 
                    !grepl("ftime", rownames(foo)) &
                    grepl(":1", rownames(foo)), 
                1:2]
table_cb <- cbind(table_cb[,1], 
                  table_cb[,1] - z_value * table_cb[,2],
                  table_cb[,1] + z_value * table_cb[,2])
table_cb <- round(exp(table_cb), 2)

table_cox <- round(summary(model_cox)$conf.int[,-2], 2)
rownames(table_cb) <- rownames(table_cox)
colnames(table_cb) <- colnames(table_cox)

@@

table_cb %>% 
  dplyr::inner_join(table_cox, by = "Covariates") %>%
  dplyr::mutate(Covariates = dplyr::case_when(
    Covariates == "SexM" ~ "Sex",
    Covariates == "DAML" ~ "Disease",
    Covariates == "PhaseCR2" ~ "Phase (CR2 vs. CR1)",
    Covariates == "PhaseCR3" ~ "Phase (CR3 vs. CR1)",
    Covariates == "PhaseRelapse" ~ "Phase (Relapse vs. CR1)",
    Covariates == "SourcePB" ~ "Source",
    TRUE ~ Covariates
    )) %>% 
  knitr::kable(format = "latex", booktabs = TRUE,
               col.names = c("Covariates", "HR", "95% CI", "HR", "95% CI"),
               caption = "Estimates and confidence intervals for the hazard ratios for each coefficient. Both estimates from case-base sampling and Cox regression are presented.") %>% 
  kable_styling() %>% 
  add_header_above(c(" " = 1, "Case-Base" = 2, "Cox" = 2))
```

@@

# Data.frame containing risk profile
newdata <- data.frame("Sex" = factor(c("F", "F"), 
                                     levels = levels(bmtcrr[,"Sex"])),
                      "D" = c("ALL", "AML"), # Both diseases
                      "Phase" = factor(c("Relapse", "Relapse"), 
                                       levels = levels(bmtcrr[,"Phase"])),
                      "Age" = c(35, 35),
                      "Source" = factor(c("PB", "PB"), 
                                        levels = levels(bmtcrr[,"Source"])))

# Estimate absolute risk curve
risk_cb <- absoluteRisk(object = model_cb, time = time_points,
                        method = "numerical", newdata = newdata)
```

@@

```{r fg_risk, eval = eval_cs2, echo = TRUE}
library(timereg)
model_fg <- comp.risk(Event(ftime, Status) ~ const(Sex) + const(D) + 
                          const(Phase) + const(Source) + const(Age), 
                      data = bmtcrr, cause = 1, model = "fg")

@@

```{r bmtcrr-risk, echo = FALSE, fig.cap="\\label{fig:compAbsrisk} Absolute risk curve for a fixed covariate profile and the two disease groups. The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.", eval = eval_cs2}
library(ggplot2)
risk_all <- dplyr::bind_rows(
    data.frame(Time = time_points, 
               Method = "Case-base", 
               Risk = risk_cb[,2], 
               Disease = "ALL",
               stringsAsFactors = FALSE),
    data.frame(Time = time_points, 
               Method = "Case-base", 
               Risk = risk_cb[,3], 
               Disease = "AML",
               stringsAsFactors = FALSE),
    data.frame(Time = time_points,
               Method = "Fine-Gray",
               Risk = risk_fg$P1[1,],
               Disease = "ALL",
               stringsAsFactors = FALSE),
    data.frame(Time = time_points,
               Method = "Fine-Gray",
               Risk = risk_fg$P1[2,],
               Disease = "AML",
               stringsAsFactors = FALSE)
    )

ggplot(risk_all, aes(x = Time, y = Risk, colour = Method)) + 
  # geom_line for smooth curve
  geom_line(data = dplyr::filter(risk_all, Method == "Case-base")) + 
  # geom_step for step function
  geom_step(data = dplyr::filter(risk_all, Method != "Case-base")) + 
  facet_grid(Disease ~ .) +
  ylim(c(0,1)) +
  theme_minimal() +
  theme(legend.position = 'top') +    
  xlab("Time (in Months)") + ylab("Relapse risk")
```

@@

```{r support-poptime, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:support-poptime} Population-time plot for the SUPPORT data."}

completeData<-casebase::support
workingCompleteData=model.matrix(death~ .-d.time,data=completeData)[,-c(1)]#remove intercept
#Create u and xCox, which will be used when fitting Cox with glmnet.
newData=workingCompleteData
x=workingCompleteData
#x and y will be used to fit the casebase model under glmnet.
#x=as.matrix(sparse.model.matrix(death~ .-d.time+0,data=workingData))
y=data.matrix(completeData[,c(2,5)])

support_popTime <- popTime(support,
                           time = "d.time", event = "death",percentile_number = 0.31)
plot(support_popTime)
```

>>>>

---
documentclass: jss
author:
  - name: Sahir Bhatnagar *
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{sahir.bhatnagar@mail.mcgill.ca}
    url: http://sahirbhatnagar.com/
  - name: Maxime Turgeon *
    affiliation: 'University of Manitoba \AND'
    address: >
      186 Dysart Road
      Winnipeg, MB, Canada R3T 2N2
    email: \email{max.turgeon@umanitoba.ca}
    url: https://maxturgeon.ca/
  - name: Jesse Islam 
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{jesse.islam@mail.mcgill.ca}
  - name: James Hanley
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{james.hanley@mcgill.ca}
    url: http://www.medicine.mcgill.ca/epidemiology/hanley/
  - name: Olli Saarela
    affiliation: University of Toronto
    address: >
      Dalla Lana School of Public Health, 
      155 College Street, 6th floor, 
      Toronto, Ontario 
      M5T 3M7, Canada
    email: \email{olli.saarela@utoronto.ca}
    url: http://individual.utoronto.ca/osaarela/
title:
  formatted: "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
  # If you use tex in the formatted title, also supply version without
  plain:     "casebase: An Alternative Framework For Survival Analysis"
  # For running headers, if needed
  short:     "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
abstract: >
  The abstract of the article. * joint co-authors
keywords:
  # at least one keyword must be supplied
  formatted: [keywords, not capitalized, "\\proglang{Java}"]
  plain:     [keywords, not capitalized, Java]
preamble: >
  \usepackage{amsmath}
  \usepackage{longtable}
  \usepackage{graphicx}
  \usepackage{tabularx}
  \usepackage{float}
  \usepackage{booktabs}
  \usepackage{makecell}
  \DeclareUnicodeCharacter{2500}{-}
output: 
  rticles::jss_article:
    number_sections: TRUE     #added argument option 
    citation_package: "natbib"  #All my citations use biblatex, not natbib. 
biblio-style: jss      #Listed to use in JSS Instructions for Authors, but not in template by default. 
bibliography: references.bib  #Also not included in template by default. 
fig_width: 12
fig_height: 9
fig_caption: true
---

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = TRUE,
  comment = "#>",
  fig.path = "../figures/"
)
options(kableExtra.latex.load_packages = FALSE)
library(tidyverse)
library(magrittr)
library(survival)
library(casebase)
# devtools::load_all('../../../casebase')#testing
library(splines)
library(tibble)
library(glmnet)
library(data.table)
library(kableExtra)
library(pracma)
options(digits = 2, scipen = 999)
# set the seed for reproducible output
set.seed(1234)

eval_introduction <- FALSE
eval_theory <- FALSE
eval_implementation <- FALSE
eval_cs1 <- FALSE
eval_cs2 <- FALSE
eval_cs3 <- TRUE
eval_cs4 <- FALSE
```

@@

```{r erspc-data, eval = eval_cs1}
data(ERSPC)
ERSPC$ScrArm <- factor(ERSPC$ScrArm,
  levels = c(0, 1),
  labels = c("Control group", "Screening group")
)
```

@@

```{r plot-stratified-erspc-data, eval = eval_cs1}
pt_object_strat <- casebase::popTime(ERSPC,
  event = "DeadOfPrCa",
  exposure = "ScrArm"
)
plot(pt_object_strat)
```

@@

```{r erspc-cox, eval = eval_cs1}
cox_model <- survival::coxph(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm,
  data = ERSPC
)
summary(cox_model)
```

@@

```{r erspc-cox-cif, eval = eval_cs1}
new_data <- data.frame(
  ScrArm = c("Control group", "Screening group"),
  ignore = 99
)

plot(survfit(cox_model, newdata = new_data),
  xlab = "Years since Randomization",
  ylab = "Cumulative Incidence (%)",
  fun = "event",
  xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
  main = sprintf(
    "Estimated Cumulative Incidence (risk) of Death from Prostate 
                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
    exp(coef(cox_model)),
    exp(confint(cox_model))[1],
    exp(confint(cox_model))[2]
  ),
  ylim = c(0, 2), yscale = 100
)
legend("topleft",
  legend = c("Control group", "Screening group"),
  col = c("red", "blue"),
  lty = c(1, 1),
  bg = "gray90"
)
```

@@

```{r erspc-casebase-exponential, eval = eval_cs1}
casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm,
  data = ERSPC,
  ratio = 100
)

summary(casebase_exponential)
exp(coef(casebase_exponential)[2])
exp(confint(casebase_exponential)[2, ])
```

@@

```{r erspc-casebase-cif, eval = eval_cs1}
smooth_risk_exp <- casebase::absoluteRisk(
  object = casebase_exponential,
  time = seq(0, 15, 0.1),
  newdata = new_data
)

plot(survfit(cox_model, newdata = new_data),
  xlab = "Years since Randomization",
  ylab = "Cumulative Incidence (%)",
  fun = "event",
  xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
  main = sprintf(
    "Estimated Cumulative Incidence (risk) of Death from Prostate 
                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
    exp(coef(cox_model)),
    exp(confint(cox_model))[1],
    exp(confint(cox_model))[2]
  ),
  ylim = c(0, 2), yscale = 100
)
lines(smooth_risk_exp[, 1], smooth_risk_exp[, 2], col = "red", lty = 2)
lines(smooth_risk_exp[, 1], smooth_risk_exp[, 3], col = "blue", lty = 2)

legend("topleft",
  legend = c(
    "Control group (Cox)", "Control group (Casebase)",
    "Screening group (Cox)", "Screening group (Casebase)"
  ),
  col = c("red", "red", "blue", "blue"),
  lty = c(1, 2, 1, 2),
  bg = "gray90"
)
```

@@

```{r erspc-casebase-weibull, eval = eval_cs1}
casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm,
  data = ERSPC,
  ratio = 100
)

@@

```{r erspc-casebase-weibull-cif, eval = eval_cs1}
smooth_risk_time <- casebase::absoluteRisk(
  object = casebase_time,
  time = seq(0, 15, 0.1),
  newdata = new_data
)

plot(survfit(cox_model, newdata = new_data),
  xlab = "Years since Randomization",
  ylab = "Cumulative Incidence (%)",
  fun = "event",
  xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
  main = sprintf(
    "Estimated Cumulative Incidence (risk) of Death from Prostate 
                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
    exp(coef(cox_model)),
    exp(confint(cox_model))[1],
    exp(confint(cox_model))[2]
  ),
  ylim = c(0, 2), yscale = 100
)
lines(smooth_risk_time[, 1], smooth_risk_time[, 2], col = "red", lty = 2)
lines(smooth_risk_time[, 1], smooth_risk_time[, 3], col = "blue", lty = 2)

legend("topleft",
  legend = c(
    "Control group (Cox)", "Control group (Casebase)",
    "Screening group (Cox)", "Screening group (Casebase)"
  ),
  col = c("red", "red", "blue", "blue"),
  lty = c(1, 2, 1, 2),
  bg = "gray90"
)
```

@@

```{r erspc-casebase-splines, eval = eval_cs1}
casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm,
  data = ERSPC,
  ratio = 100
)

@@

```{r erspc-casebase-splines-cif, eval = eval_cs1}
smooth_risk_splines <- absoluteRisk(
  object = casebase_splines,
  time = seq(0, 15, 0.1),
  newdata = new_data
)

plot(survfit(cox_model, newdata = new_data),
  xlab = "Years since Randomization",
  ylab = "Cumulative Incidence (%)",
  fun = "event",
  xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
  main = sprintf(
    "Estimated Cumulative Incidence (risk) of Death from Prostate 
                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
    exp(coef(cox_model)),
    exp(confint(cox_model))[1],
    exp(confint(cox_model))[2]
  ),
  ylim = c(0, 2), yscale = 100
)
lines(smooth_risk_splines[, 1], smooth_risk_splines[, 2], col = "red", lty = 2)
lines(smooth_risk_splines[, 1], smooth_risk_splines[, 3], col = "blue", lty = 2)

legend("topleft",
  legend = c(
    "Control group (Cox)", "Control group (Casebase)",
    "Screening group (Cox)", "Screening group (Casebase)"
  ),
  col = c("red", "red", "blue", "blue"),
  lty = c(1, 2, 1, 2),
  bg = "gray90"
)
```

@@

```{r erspc-95ci, echo = FALSE, eval = eval_cs1}
CI_cox <- c(exp(coef(cox_model)), exp(confint(cox_model)))
CI_exp <- c(exp(coef(casebase_exponential))["ScrArmScreening group"], exp(confint(casebase_exponential))["ScrArmScreening group", ])
CI_time <- c(exp(coef(casebase_time))["ScrArmScreening group"], exp(confint(casebase_time))["ScrArmScreening group", ])
CI_splines <- c(exp(coef(casebase_splines))["ScrArmScreening group"], exp(confint(casebase_splines))["ScrArmScreening group", ])

@@

data.frame(Model = c("Cox", "Exponential", "Weibull", "Splines")) %>%
  bind_cols(as.data.frame(table_ci)) %>%
  mutate(
    CI = paste0(
      "(", round(LB, 2), ", ",
      round(UB, 2), ")"
    ),
    Width = UB - LB
  ) %>%
  dplyr::select(-LB, -UB) %>%
  knitr::kable(format = "latex", booktabs = TRUE) %>%
  kable_styling()
```

@@

```{r erscp-aic, eval = eval_cs1}
c(
  "Exp." = AIC(casebase_exponential),
  "Weib." = AIC(casebase_time),
  "Splines" = AIC(casebase_splines)
)
```

@@

```{r echo = FALSE, eval = eval_cs2}
tibble::tribble(
  ~Variable, ~Description, ~`Statistical summary`,
  "\\code{Sex}", "Sex", "M=Male (100)",
  "", "", "F=Female (77)",
  "\\code{D}", "Disease", "ALL (73)",
  "", "", "AML (104)",
  "\\code{Phase}", "Phase", "CR1 (47)",
  "", "", "CR2 (45)",
  "", "", "CR3 (12)",
  "", "", "Relapse (73)",
  "\\code{Source}", "Type of transplant", "BM+PB (21)",
  "", "", "PB (156)",
  "\\code{Age}", "Age of patient (years)", "4--62",
  "", "", "30.47 (13.04)",
  "\\code{ftime}", "Failure time (months)", "0.13--131.77",
  "", "", "20.28 (30.78)",
  "\\code{Status}", "Status indicator", "0=censored (46)",
  "", "", "1=relapse (56)",
  "", "", "2=competing event (75)"
) %>%
  knitr::kable(
    format = "latex", booktabs = TRUE,
    caption = "Baseline characteristics of patients in the stem-cell transplant study.",
    label = "table1bmtcrr",
    escape = FALSE
  ) %>%
  kable_styling()
```

@@

pt_object_comp <- bmtcrr %>%
  mutate(Status = case_when(
    Status == 1 ~ 2L,
    Status == 2 ~ 1L,
    TRUE ~ Status
  )) %>%
  casebase::popTime(event = "Status", time = "ftime")

plot_grid(plot(pt_object) + ggtitle("Relapse"),
  plot(pt_object_comp, point.colour = "blue") +
    ggtitle("Competing events"),
  labels = c("A", "B")
)
```

@@

```{r bmtcrr-casebase-weibull, warning = FALSE, eval = eval_cs2}
model_cb <- fitSmoothHazard(
  Status ~ ftime + Sex + D + Phase + Source + Age,
  data = bmtcrr,
  ratio = 100,
  time = "ftime"
)
```

@@

```{r bmtcrr-cox, echo = TRUE, eval = eval_cs2}
library(survival)
# Treat competing event as censoring
model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
  data = bmtcrr
)
```

```{r bmtcrr-cis, echo=FALSE, eval = eval_cs2}
# Table of coefficients----
library(glue)
z_value <- qnorm(0.975)
foo <- summary(model_cb)@coef3
table_cb <- foo[
  !grepl("Intercept", rownames(foo)) &
    !grepl("ftime", rownames(foo)) &
    grepl(":1", rownames(foo)),
  1:2
]
table_cb <- cbind(
  table_cb[, 1],
  table_cb[, 1] - z_value * table_cb[, 2],
  table_cb[, 1] + z_value * table_cb[, 2]
)
table_cb <- round(exp(table_cb), 2)

table_cox <- round(summary(model_cox)$conf.int[, -2], 2)
rownames(table_cb) <- rownames(table_cox)
colnames(table_cb) <- colnames(table_cox)

@@

table_cb %>%
  dplyr::inner_join(table_cox, by = "Covariates") %>%
  dplyr::mutate(Covariates = dplyr::case_when(
    Covariates == "SexM" ~ "Sex",
    Covariates == "DAML" ~ "Disease",
    Covariates == "PhaseCR2" ~ "Phase (CR2 vs. CR1)",
    Covariates == "PhaseCR3" ~ "Phase (CR3 vs. CR1)",
    Covariates == "PhaseRelapse" ~ "Phase (Relapse vs. CR1)",
    Covariates == "SourcePB" ~ "Source",
    TRUE ~ Covariates
  )) %>%
  knitr::kable(
    format = "latex", booktabs = TRUE,
    col.names = c("Covariates", "HR", "95% CI", "HR", "95% CI"),
    caption = "Estimates and confidence intervals for the hazard ratios for each coefficient. Both estimates from case-base sampling and Cox regression are presented."
  ) %>%
  kable_styling() %>%
  add_header_above(c(" " = 1, "Case-Base" = 2, "Cox" = 2))
```

@@

# Data.frame containing risk profile
newdata <- data.frame(
  "Sex" = factor(c("F", "F"),
    levels = levels(bmtcrr[, "Sex"])
  ),
  "D" = c("ALL", "AML"), # Both diseases
  "Phase" = factor(c("Relapse", "Relapse"),
    levels = levels(bmtcrr[, "Phase"])
  ),
  "Age" = c(35, 35),
  "Source" = factor(c("PB", "PB"),
    levels = levels(bmtcrr[, "Source"])
  )
)

# Estimate absolute risk curve
risk_cb <- absoluteRisk(
  object = model_cb, time = time_points,
  method = "numerical", newdata = newdata
)
```

@@

```{r fg_risk, eval = eval_cs2, echo = TRUE}
library(timereg)
model_fg <- comp.risk(Event(ftime, Status) ~ const(Sex) + const(D) +
  const(Phase) + const(Source) + const(Age),
data = bmtcrr, cause = 1, model = "fg"
)

@@

```{r bmtcrr-risk, echo = FALSE, fig.cap="\\label{fig:compAbsrisk} Absolute risk curve for a fixed covariate profile and the two disease groups. The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.", eval = eval_cs2}
library(ggplot2)
risk_all <- dplyr::bind_rows(
  data.frame(
    Time = time_points,
    Method = "Case-base",
    Risk = risk_cb[, 2],
    Disease = "ALL",
    stringsAsFactors = FALSE
  ),
  data.frame(
    Time = time_points,
    Method = "Case-base",
    Risk = risk_cb[, 3],
    Disease = "AML",
    stringsAsFactors = FALSE
  ),
  data.frame(
    Time = time_points,
    Method = "Fine-Gray",
    Risk = risk_fg$P1[1, ],
    Disease = "ALL",
    stringsAsFactors = FALSE
  ),
  data.frame(
    Time = time_points,
    Method = "Fine-Gray",
    Risk = risk_fg$P1[2, ],
    Disease = "AML",
    stringsAsFactors = FALSE
  )
)

ggplot(risk_all, aes(x = Time, y = Risk, colour = Method)) +
  # geom_line for smooth curve
  geom_line(data = dplyr::filter(risk_all, Method == "Case-base")) +
  # geom_step for step function
  geom_step(data = dplyr::filter(risk_all, Method != "Case-base")) +
  facet_grid(Disease ~ .) +
  ylim(c(0, 1)) +
  theme_minimal() +
  theme(legend.position = "top") +
  xlab("Time (in Months)") +
  ylab("Relapse risk")
```

@@

```{r support-poptime, echo=FALSE, eval = eval_cs3, fig.cap="\\label{fig:support-poptime} Population-time plot for the SUPPORT data."}
support <- casebase::support

support_popTime <- popTime(support,
  time = "d.time", event = "death", percentile_number = 0.31
)
plot(support_popTime)
```
